---
title: Mouse ileum cell interactions
vignette: >
  % \VignetteIndexEntry{Mouse ileum cell interactions}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r, message = FALSE, warning = FALSE}
library(MerfishData)
library(ExperimentHub)
library(ggplot2)
library(imcRtools)
```

# Data retrieval

```{r ehub}
eh <- ExperimentHub()
query(eh, c("MerfishData", "ileum"))
```

```{r spe-ileum, message = FALSE}
spe <- MerfishData::MouseIleumPetukhov2021(segmentation = "baysor",
                                           use.images = FALSE,
                                           use.polygons = FALSE)
spe
```

# Spatial proximity analysis

## Construct spatial graph 

To identify cells that are in close proximity to each other, we use the
`buildSpatialGraph` function from the
[imcRtools](https://bioconductor.org/packages/imcRtools) package.
The function constructs interaction graphs using either (i) cell-centroid expansion,
(ii) _k_-nearest neighbor search, or (iii) delaunay triangulation.

Here, we use a k-nearest neighbor search with k = 5, which accordingly constructs
a graph based on the 5 closest cells for each cell.

```{r}
spe <- imcRtools::buildSpatialGraph(spe,
                                    img_id = "sample_id",
                                    type = "knn",
                                    k = 5, 
                                    coords = c("x", "y")) 
colPairs(spe)
colPair(spe, "knn_interaction_graph")
```

## Neighborhood permutation testing

To identify recurrent interactions between cell type labels, we use the
`countInteractions` and `testInteractions` functions from the
[imcRtools](https://bioconductor.org/packages/imcRtools) package.

The `countInteractions` function counts the number of edges (interactions)
between each pair of cell labels.
Simplified, it counts for each cell of type A the number of neighbors of type B.

This count can be averaged within each grouping level (e.g. image/sample) in different ways.
Here, we use `method = "classic"` to divide the interaction count is divided by
the total number of cells of type A. The final count can be interpreted as
"How many neighbors of type B does a cell of type A have on average?". 

```{r}
ia.counts <- imcRtools::countInteractions(spe,
                                          group_by = "sample_id",
                                          label = "leiden_final",
                                          colPairName = "knn_interaction_graph",
                                          method = "classic")
ind <- order(ia.counts$ct, decreasing = TRUE)
ia.counts <- ia.counts[ind,]
ia.counts
```

Given these average interaction counts between cell type labels, we can then test
if the obtained count is larger or smaller compared to what is expected from a
random distribution of cell labels. 

For this, we use the `testInteractions` function from the 
[imcRtools](https://bioconductor.org/packages/imcRtools) package, which permutes
the cell labels and counts interactions as described above. 
This approach generates a distribution of the interaction count under a random
distribution of cell labels. 
The observed interaction count is compared against this null distribution to
derive empirical p-values:

```{r}
ia.sig <- imcRtools::testInteractions(spe,
                                      group_by = "sample_id",
                                      label = "leiden_final",
                                      colPairName = "knn_interaction_graph",
                                      method = "classic")
sort.df <- data.frame(p = ia.sig$p, ct = -ia.sig$ct)
ind <- do.call(order, sort.df)
ia.sig <- ia.sig[ind,]
ia.sig
```

This seem to primarily work out that cells of the same type tend to be close to
each other, which is also apparent from spatial domain organization in the image.

Which cells of different cell types tend to be disproportionately often close to
each other?

```{r}
ind <- ia.sig$from_label != ia.sig$to_label
head(ia.sig[ind,], n = 10)
```

Let's also visualize the pairwise interaction counts between cell type labels:

```{r}
df <- subset(ia.sig, ct > 0)
df <- data.frame(df)
df$sig <- factor(df$sig, levels = c(TRUE, FALSE))
df$nlogp <- -log10(df$p)
```

Here we only look at whether an interaction count surpassed a signficance threshold
of 0.01.

```{r, fig.width = 10, fig.height = 10}
ggplot(df, aes(x = from_label, y = to_label, size = ct, color = sig)) + 
    geom_point() +
    xlab("from") + 
    ylab("to") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Here we also look at the magnitude of the effect.

```{r, fig.width = 10, fig.height = 10}
ggplot(df, aes(x = from_label, y = to_label, size = ct, color = nlogp)) + 
    geom_point() + 
    xlab("from") + 
    ylab("to") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Ligand-receptor co-expression analysis


# Ligand-target co-expression analysis


