---
title: Mouse ileum morphological features
vignette: >
  % \VignetteIndexEntry{Mouse ileum morphological features}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r, message = FALSE}
library(MerfishData)
library(ExperimentHub)
library(ggpubr)
library(reshape2)
```

# Data retrieval

```{r ehub}
eh <- ExperimentHub()
query(eh, c("MerfishData", "ileum"))
```

```{r spe-ileum, message = FALSE}
spe <- MerfishData::MouseIleumPetukhov2021(segmentation = "baysor",
                                           use.images = FALSE,
                                           use.polygons = FALSE)
spe
```

Here, we inspect the available basic morphological features from the Baysor 
segmentation:  

```{r cdat}
colData(spe)
```

# Number of transcripts

We stratify the available cell features by cell type, starting with the
number of transcripts expressed in each cell:

```{r n-transcripts}
spl <- split(spe$n_transcripts, spe$leiden_final)
df <- reshape2::melt(spl)
colnames(df) <- c("n_transcripts", "leiden_final")
```

We then plot the distribution of number of transcripts by cell type:

```{r n-transcripts-plot, fig.width = 6, fig.height = 6}
bp <- ggpubr::ggboxplot(df, x = "leiden_final", y = "n_transcripts",
                        fill = "leiden_final", ggtheme = theme_bw(),
                        palette = "ucscgb", legend = "none")
bp + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From this plot it is apparent that, for example, T (CD8+) cells express more
transcripts than T (CD4+) cells, which we can corroborate with a simple t-test:

```{r}
t.test(spl[["T (CD8+)"]], spl[["T (CD4+)"]])
```

(This is overdispersed count data though):

```{r}
mean(spl[["T (CD8+)"]])
var(spl[["T (CD8+)"]])
length(spl[["T (CD8+)"]])
```

# Density

```{r n-density}
spl <- split(spe$density, spe$leiden_final)
df <- reshape2::melt(spl)
colnames(df) <- c("density", "leiden_final")
df <- subset(df, leiden_final != "Removed")
```

We then plot the distribution of by cell type:

```{r n-density-plot, fig.width = 6, fig.height = 6}
bp <- ggpubr::ggboxplot(df, x = "leiden_final", y = "density",
                        fill = "leiden_final", ggtheme = theme_bw(),
                        palette = "ucscgb", legend = "none")
bp + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Elongation

```{r n-elong}
spl <- split(spe$elongation, spe$leiden_final)
df <- reshape2::melt(spl)
colnames(df) <- c("elongation", "leiden_final")
df <- subset(df, leiden_final != "Removed")
```

We then plot the distribution of by cell type:

```{r n-elong-plot, fig.width = 6, fig.height = 6}
bp <- ggpubr::ggboxplot(df, x = "leiden_final", y = "elongation",
                        fill = "leiden_final", ggtheme = theme_bw(),
                        palette = "ucscgb", legend = "none")
bp + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Area

```{r n-area}
spl <- split(spe$area, spe$leiden_final)
df <- reshape2::melt(spl)
colnames(df) <- c("area", "leiden_final")
df <- subset(df, leiden_final != "Removed")
```

We then plot the distribution of by cell type:

```{r n-area-plot, fig.width = 6, fig.height = 6}
bp <- ggpubr::ggboxplot(df, x = "leiden_final", y = "area",
                        fill = "leiden_final", ggtheme = theme_bw(),
                        palette = "ucscgb", legend = "none")
bp + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Confidence

```{r n-conf}
spl <- split(spe$avg_confidence, spe$leiden_final)
df <- reshape2::melt(spl)
colnames(df) <- c("avg_confidence", "leiden_final")
df <- subset(df, leiden_final != "Removed")
```

We then plot the distribution of by cell type:

```{r n-conf-plot, fig.width = 6, fig.height = 6}
bp <- ggpubr::ggboxplot(df, x = "leiden_final", y = "avg_confidence",
                        fill = "leiden_final", ggtheme = theme_bw(),
                        palette = "ucscgb", legend = "none")
bp + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
