---
title: Mouse ileum segmentation
vignette: >
  % \VignetteIndexEntry{Mouse ileum segmentation}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r, message = FALSE, warning = FALSE}
library(MerfishData)
library(ExperimentHub)
library(alphahull)
library(ggplot2)
library(sp)
```

# Data retrieval

```{r ehub}
eh <- ExperimentHub()
query(eh, c("MerfishData", "ileum"))
```

```{r spe-ileum, message = FALSE}
spe <- MerfishData::MouseIleumPetukhov2021(segmentation = "baysor",
                                           use.images = FALSE)
spe
```

# Polygon segmentation borders

Here, we inspect the polygon segmentation borders obtained from the Baysor
segmentation:  

```{r poly}
poly <- metadata(spe)$polygons
head(poly)
```

We first check how many cells / polygons do we have for each z-layer:

```{r poly-per-z}
spl <- split(poly[,"cell"], poly[,"z"])
spl <- lapply(spl, unique)
lengths(spl)
```

That means we have 3866 cells in the first z-plane.

Question: how to relate the polygons to the 5,800 cells that we have in the
segmentation counts matrix? 

Answer: we can use functionality from the 
[sp](https://cran.r-project.org/web/packages/sp/index.html)
package for assigning cell centroids to polygons they are spatially contained in. 

This can be done using either the `point.in.polygon` function.
For example, we identify here cell centroid coordinates that overlap with
the first polygon of the first z-plane:

```{r point-in-poly}
poly1 <- poly[poly[,"cell"] == 1 & poly[,"z"] == 1, c("x", "y")]
res <- sp::point.in.polygon(spatialCoords(spe)[,"x"],
                            spatialCoords(spe)[,"y"],
                            poly1[,"x"],
                            poly1[,"y"])
head(res)
table(res)
```

This means that the first cell in the assay corresponds to the first polygon in
the first z-layer.

Or we can use the `over` function, which allows to more easily iterate over multiple
polygons.
For example, we identify here cell centroid coordinates that overlap with
the first two polygons of the first z-plane:

```{r}
pts <- sp::SpatialPoints(spatialCoords(spe))
p1 <- sp::Polygon(poly1[,c("x", "y")])
poly2 <- poly[poly[,"cell"] == 2 & poly[,"z"] == 1, c("x", "y")]
p2 <- sp::Polygon(poly2[,c("x", "y")])
pol <- sp::SpatialPolygons(list(sp::Polygons(list(p1), ID = "1"),
                                sp::Polygons(list(p2), ID = "2")))
```

```{r points-over-polys}
res <- sp::over(pts, pol)
head(res)
table(res)
resl <- sp::over(pts, pol, returnList = TRUE)
head(resl)
```

This means that the first two cell in the assay corresponds to the first two polygons
in the first z-layer.

Ok, so now let's do this for all cell centroids of the Baysor segmentation to
find the associated polygons:

```{r all-points-over-polys}
zspl <- split(data.frame(poly[,c("cell", "x", "y")]), poly[,"z"])
cellspl <- lapply(zspl, function(s) split(s[,c("x", "y")], s[,"cell"]))
spl <- do.call(c, cellspl)
```

We define a little helper function for creating polygon objects:

```{r create-poly}
.createPolygon <- function(i)
{
    p <- sp::Polygon(spl[[i]])
    p <- sp::Polygons(list(p), ID = as.character(i))
    return(p)
}
```

```{r create-all-polys}
pz <- lapply(seq_along(spl), .createPolygon)
pz <- sp::SpatialPolygons(pz)
res <- sp::over(pts, pz)
head(res)
resl <- sp::over(pts, pz, returnList = TRUE)
head(resl)
table(lengths(resl))
```

That means that we have 8 cells in the segmentation matrix without overlapping
polygons, and, on the other hand, 3 cells with up to 11 overlapping polygons.

As there is thus indeed no 1:1 mapping, and given that the Baysor segmentation
results are 2D, we could eg. just take the first polygon for each cell,
irrespective of the z-layer:

```{r first-poly-per-cell}
poly.per.cell <- vapply(resl, `[`, integer(1), x = 1)
head(poly.per.cell)
```

# Convex hull segmentation borders 

Alternatively, we can compute and visualize simple convex hulls instead:

```{r }
x <- BumpyMatrix::unsplitAsDataFrame(assay(spe, "molecules"))
x$column <- as.integer(x$column)
x
```

Using `base::chull`, this computes convex hulls for
5,800 cells and around 820k molecules in less than a second:

```{r}
cl <- split(data.frame(x[,c("x", "y")]), x$column)
length(cl[[1]])
system.time( clh <- lapply(cl, chull) )
clh[[1]]
length(clh[[1]])
length(clh)
```

Create data frame for plotting:

```{r}
cl.sub <- cl
for(i in seq_along(clh)) cl.sub[[i]] <- cl.sub[[i]][clh[[i]],]
clc <- do.call(rbind, cl.sub)
clc$cell <- rep(names(cl.sub), lengths(clh))
```

```{r, fig.height = 8}
clc$leiden <- spe$leiden_final[as.integer(clc$cell)]
p <- ggplot() + geom_polygon(data = clc,
                        aes(x = x, y = y, group = cell, color = leiden),
                        fill = NA)
p + theme_bw()
```

# Alpha-convex hull segmentation borders

Although convex hulls are useful for simple approximation of cell shapes given
the entirety of molecules annotated to a cell, there might also be concave 
(inward curving) elements to a cell shape.

Using the concept of an $\alpha$-convex hull as implemented in the 
[alphahull](https://cran.r-project.org/package=alphahull)
package, we are also able to reconstruct non-convex hulls from the spatial
location of each molecule annotated to a cell.

An important parameter is the shape parameter $\alpha$, which describes determines
the degree of inward curving, where the larger the $\alpha$, the more the hull
approaches a convex hull. 

Here we inspect the effect of varying $\alpha$ for the first five cells of the ileum
dataset:

```{r, fig.width = 10, fig.height = 10}
cl <- lapply(cl, unique)
par(mfrow = c(5,5))
par(pch = 20)
cols <- rainbow(8)
for(i in 1:5)
{
    for(a in seq(5, 25, by = 5))
    {
        main <- ""
        if(i == 1) main <- paste("alpha", a, sep = " = ")
        res <- alphahull::ahull(cl[[i]], alpha = a)
        plot(res, xlab = "x", ylab = "y", col = cols[i], main = main)
    }
}
```

For demonstration purpose, we compute here $\alpha$-hull for the first 8 cells
with $\alpha$ set to 20.

```{r}
ahl <- lapply(cl[1:8], alphahull::ahull, alpha = 20)
```

Plot:

```{r, fig.width = 8, fig.height = 8}
par(mfrow = c(1,1)) 
plot(ahl[[1]], col = cols[1],
     xlim = c(2000,2300), ylim = c(0,250),
     xlab = "x", ylab = "y")
for(i in 2:8)
    plot(ahl[[i]], col = cols[i], add = TRUE)
```
